---
title: "Notebook Exercises Chapter 13"
author: "Max Beauchez"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE)
library(rethinking)
```

**13E1.** Which of the following priors will produce more *shrinkage* in the estimates? (a) $\alpha_{TANK} \sim Normal(0, 1)$; (b) $\alpha_{TANK} \sim Normal(0, 2)$.

The first one, as it has a smaller prior for the standard deviation.

**13E2.** Rewrite the following model as a multilevel model.

$y_{i} \sim Binomial(1, p_{i})$

$logit(p_{i}) = \alpha_{GROUP[i]} + \beta x_{i}$

$\alpha_{GROUP} \sim Normal(0, 1.5)$

$\beta \sim Normal(0, 0.5)$

To turn this model in to multilevel model we need to make the prior for $\alpha_{GROUP}$ a function of two new parameters, $\bar{\alpha}$ and $\sigma\_{\alpha}$, each with their own hyper-priors.

$y_{i} \sim Binomial(1, p_{i})$

$logit(p_{i}) = \alpha_{GROUP[i]} + \beta x_{i}$

$\alpha_{GROUP} \sim Normal(\bar{\alpha}, \sigma_{\alpha})$

$\bar{\alpha} \sim Normal(0, 1)$

$\sigma_{\alpha} \sim Exponential(1)$

$\beta \sim Normal(0, 0.5)$

**13E3.** Rewrite the following model as a multilevel model.

$y_{i} \sim Normal(\mu_{i}, \sigma)$

$\mu_{i} = \alpha_{GROUP[i]} + \beta x_{i}$

$\alpha_{GROUP} \sim Normal(0, 5)$

$\beta \sim Normal(0, 1)$

$\sigma \sim Exponential(1)$

As before, we make the prior for $\alpha_{GROUP[i]}$ a function of two new parameters, $\bar{\alpha}$ and $\sigma_{\alpha}$, each with their own hyper-priors.

$y_{i} \sim Normal(\mu_{i}, \sigma)$

$\mu_{i} = \alpha_{GROUP[i]} + \beta x_{i}$

$\alpha_{GROUP} \sim Normal(\bar{\alpha}, \sigma_{\alpha})$

$\bar{\alpha} \sim Normal(0, 1)$

$\sigma_{\alpha} \sim Exponential(1)$

$\beta \sim Normal(0, 1)$

$\sigma \sim Exponential(1)$

**13E4.** Write a mathematical model formula for a Poisson regression with varying intercepts.

Without any knowledge of what this model is used for I can't choose any meaningful priors, so the numbers in the below definition are just placeholders.

$y_{i} \sim Poisson(\lambda_{i})$

$log\,(\lambda_{i}) = \alpha_{i} + \beta x_{i}$

$\alpha \sim Normal(\bar{\alpha}, \sigma_{\alpha})$

$\bar{\alpha} \sim Normal(0, 1)$

$\sigma_{\alpha} \sim Exponential(1)$

$\beta \sim Normal(0, 1)$

**13E5.** Write a mathematical model formula for a Poisson regression with two different kinds of varying intercepts, a cross-classified model.

As before, any numerical values are pro forma, as no information is given about what is being modelled.

$y_{i} \sim Poisson(\lambda_{i})$

$log\,(\lambda_{i}) = \alpha_{i} + \gamma_{i} + \beta x_{i}$

$\alpha \sim Normal(\bar{\alpha}, \sigma_{\alpha})$

$\bar{\alpha} \sim Normal(0, 1)$

$\sigma_{\alpha} \sim Exponential(1)$

$\gamma_{i} \sim Normal(\bar{\gamma}, \sigma_{\gamma})$

$\bar{\gamma} \sim Normal(0, 1)$

$\sigma_{\gamma} \sim Exponential(1)$

$\beta \sim Normal(0, 1)$

**13M1.** Revisit the Reed frog survival data, `data(reedfrogs)`, and add the `predation` and `size` treatment variables to the varying intercepts model. Consider models with either main effect alone, both main effects, as well as a model including both and their interaction. Instead of focusing on inferences about these two predictor variables, focus on the inferred variation across tanks. Explain why it changes as it does across models.

```{r}
data(reedfrogs)
d <- reedfrogs
```

**13M2.** Compare the models you fit just above, using WAIC. Can you reconcile the differences in WAIC with the posterior distributions of the models?

**13M3.** Re-estimate the basic Reed frog varying intercept model, but now using a Cauchy distribution in place of the Gaussian distribution for the varying intercepts. That is, fit this model:

$s_{i} \sim Binomial(n_{i}, p_{i})$

$logit(p_{i}) = \alpha_{TANK[i]}$

$\alpha_{TANK} \sim Cauchy(\alpha, \sigma)$

$\alpha \sim Normal(0, 1)$

$\sigma \sim Exponential(1)$

(You are likely to see many divergent transitions for this model. Can you figure out why? Can you fix them? Compare the posterior means of the intercepts, $\alpha_{TANK}$, to the posterior means produced in the chapter, using the customary Gaussian prior. Can you explain the pattern of differences? Take note of any change in the mean $\alpha$ as well).

**13M4.** Now use a Student-t distribution with $\nu$ = 2 for the intercepts:

$\alpha_{TANK} \sim Student(2, \alpha, \sigma)$

Refer back to the Student-t example in Chapter 7 (page 234), if necessary. Compare the resulting posterior to both the original model and the Cauchy model in 13M3. Can you explain the differences and similarities in shrinkage in terms of the properties of these distributions?

**13M5.** Modify the cross-classified chimpanzees model `m13.4` so that the adaptive prior for blocks contains a parameter $\bar{\gamma}$ for its mean:

$\gamma_{j} \sim Normal(\bar{\gamma}, \sigma_{\gamma}$ $\bar{\gamma} \sim Normal(0, 1.5)$

Compare this model to `m13.4`. What has including $\bar{\gamma}$ done?

**13M6.** Sometimes the prior and the data (through the likelihood) are in conflict, because they concentrate around different regions of parameter space. What happens in these cases depends a lot upon the shape of the tails of the distributions. Likewise, the tails of distributions can strongly influence whether or not outliers are shrunk towards the mean. I want you to consider four different models to fit to one observation at *y* = 0. The models differ only in the distributions assigned to the likelihood and prior. Here are the four models::

Model NN:

$y \sim Normal(\mu, 1)$

$\mu \sim Normal(10, 1)$

Model NT:

$y \sim Normal(\mu, 1)$

$\mu \sim Student(2, 10, 1)$

Model TN:

$y \sim Student(2, \mu, 1)$

$\mu \sim Normal(10, 1)$

Model TT:

$y \sim Student(2, \mu, 1)$

$\mu \sim Student(2, 10, 1)$

Estimate the posterior distributions for these models and compare them. Can you explain the results, using the properties of the distributions?

**13H1.** In 1980, a typical Bengali woman could have 5 or more children in her lifetime. By the year 2000, a typical Bengali woman had only 2 or 3. You're going to look at a historical set of data, when contraception was widely available but many families chose not to use it. These data reside in `data(bangladesh)` and come from the 1988 Bangladesh Fertility Survey. Each row is one of the 1934 women. There are six variables, but you can focus on two of them for this practice problem:

  1. `district`: ID number of administrative district each woman resided in
  2. `use.contraception`: An indicator (0/1) of whether the woman was using contraception
  
The first thing to do is ensure that the cluster variable, `district`, is a contiguous set of integers. Recall that these values will be index values inside the model. If there are gaps, you'll have parameters for which there is no data to inform them. Worse, the model probably won't run. Look at the unique values of the `district` variable:

```{r}
#13.40
sort(unique(d$district))
```

District 54 is absent. So `district isn't yet a good index variable, because it's not contiguous. This is easy to fix. Just make a new variable that is contiguous. This is enough to do it:

```{r}
#13.41
d$district_id <- as.integer(as.factor(d$district))
sort(unique(d$district_id))
```

Now there are 60 values, contiguous integers 1 to 60. Now, focus on predicting `use.contraception`, clustered by `district_id`. Fit both (1) a traditional fixed-effects model that uses an index variable for district and (2) a multilevel model with varying intercepts for district. Plot the predicted proportions of women in each district using contraception, for both the fixed-effects model and the varying-effect model. That is, make a plot in which district ID is on the horizontal axis and expected proportion using contraception is on the vertical. Make one plot for each model, or layer them on the same plot, as you prefer. How do the models disagree? Can you explain the pattern of disagreement? In particular, can you explain the most extreme cases of disagreement, both why they happen where they do and why the models reach different inferences?

**13H2.** Return to `data(Trolley)` from Chapter 12. Define and fit a varying intercepts model for these data. Cluster intercepts on individual participants, as indicated by the unique values in the `id` variable. Include `action`, `intention`, and `contact` as ordinary terms. Compare the varying intercepts model and a model that ignores individuals, using both WAIC and posterior predictions. What is the impact of a individual variations in these data?

**13H3.** The `Trolley` data are also clustered by `story`, which indicates a unique narrative for each vignette. Define and fit a cross-classified varying intercepts model with both `id` and `story`. Use the same ordinary terms as in the previous problem. Compare this model to the previous models. What do you infer about the impact of different stories on responses?

**13H4.** Revisit the Reed frog survival data, `data(reedfrogs)`, and add the `predation` and `size` treatment variables to the varying intercepts model. Consider models with either predictor alone, both predictors, as well as a model including their interaction. What do you infer about the causal influence of these predictor variables? Also focus on the inferred variation across tanks (the \sigma across tanks). Explain why it changes as it does across models with different predictors included.
